
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)

NORS = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/NORS_refined.csv", header = TRUE)
NOAA = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/NOAA_cleaned.csv", header = TRUE)

#keeps reading in the observation number as a column
#idk why but i'll just delete it manually
NOAA$X <- NULL
NORS$X <- NULL
```

NORS observation distributions (preliminary)
```{r}
#by year
NORS_year = NORS |> group_by(Year) |> 
  summarize(n = n())

#by year HISTOGRAM
NORS |> group_by(Year) |> 
  summarize(n = n()) |> 
  ggplot(aes(Year))+
  geom_histogram(aes(weight = n))

NORS_state = NORS |> group_by(State) |> 
  summarize(n = n())

#tbh i'm not even sure how to go about cleaning this
#what even are relevant categories to divide by?

#no decent data until 1998
#data also ends at 2022 but we will cut it off at 2019 to avoid influence from the pandemic

NORS_cropped = NORS |> 
  filter(Year >= 1998 & Year <= 2019)
```

NORS (updated)
```{r}
#NEW by year HISTOGRAM
NORS_cropped |> group_by(Year) |> 
  summarize(n = n()) |> 
  ggplot(aes(Year))+
  geom_histogram(aes(weight = n), bins = 21)


#the amount of reporting is still definitely skewed towards more recent years
#but we're up to an n factor that I'm not too worried about it?

```

NOAA observations distributions
```{r}
#we've already cropped some of the older data from the NOAA dataset
#but let's be sure we're only looking at 1998-2019
#starting with 2646 observations in the original dataframe

NOAA_cropped = drop_na(NOAA, c("Oct", "Nov", "Dec"))
#this drops 98 entries (2548 left)
#HOWEVER: please note that just filtering by year also drops these entries
#so the NAs aren't even in our scope and we have full data for the entire timespan

NOAA_cropped = NOAA_cropped |> 
  filter(Year >= 1998 & Year <= 2019)
#2156 entries left (dropped 490)

```


run this to delete all objects except the most recent datasets
```{r}
#NORS variables
rm(list = c("NORS", "NORS_state", "NORS_year"))

#NOAA variables
rm(list = c("NOAA"))

```

The NOAA and the NORS data have now been restricted to the same time frame.
Moving onto summary stats
NOAA
- temperature range
  - change over the decades
  - variation within a year
- state coverage
  - regions?
  
NORS
- disease frequency over time
  - can we even do this with varying levels of de
  

NOAA summary stats
```{r}



```





exploring lists (and counts) of each unique etiology entry to prep for pruning
```{r}

single_factors = refined |>  
  filter(!is.na(Etiology)) |> 
  filter(Etiology != "Multiple")|> 
  group_by(Etiology) |> 
  summarize(n = n())

single_factors
#listed_factors = refined |> 
#  filter(Etiology == "Multiple") |> 
#  group_by(Etiology.Expanded) |> 
#  summarize(n = n())

write.csv(single_factors, "etiologies_single.csv")
#write.csv(listed_factors, "listed_factors.csv")

```


```

