## CLEANING
goals
- fix variables
- fix primary.mode into categorical & rename to make it easier
- split etiology into 2 variables
- 1 is etiology (clean)
- 2 is etiology (expanded)


##NOTE: this code uses the new R pipe ( |> ) so rstudio should be updated to 4.1+

load in the database
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)

raw = read.csv("/Users/machaango/OneDrive/Desktop/thesis/csv_storage/NORSdata.csv", header = TRUE)
```

tidying Primary.Mode Variable
```{r}
#make variable categorical and rename categories to remove spaces (makes them easier to call)
raw$Primary.Mode = as.factor(raw$Primary.Mode)
raw$Primary.Mode = fct_recode(raw$Primary.Mode, "Animal" = "Animal contact", "Interpersonal" = "Person-to-person", "Unknown" = "Indeterminate/unknown", "Environmental" = "Environmental contamination other than food/water")
#REMINDER: "Environmental" group does NOT include water or food borne infections

message("done!")
```

prep work for etiology and cleaning columns
```{r}
#create new column in preparation for cleaning the etiology variable
#move it to the right spot & drop unneeded columns
#also drops values w/unknown or NA etiologys
refined = raw |>  
  mutate(Etiology.Expanded = Etiology) |>  
  relocate(Etiology.Expanded, .after = Etiology) |>  
  select(!c(Food.Contaminated.Ingredient, Animal.Type, Water.Type, Water.Exposure, IFSAC.Category, Food.Vehicle, Serotype.or.Genotype))


#can we move this down to below the cleaning loop?
dropped = refined |> 
  filter((Etiology == "Unknown") | (Etiology == "")) |> 
  mutate(Etiology = na_if(Etiology, ""))

refined = refined |> 
  mutate(Etiology = na_if(Etiology,  "")) |> 
  drop_na(Etiology) |> 
  filter(Etiology != "Unknown")


message("done!")
```


tidying scrap chunk (code to be moved to labeled chunk once workshopped)

cutting out
- variants of the same disease (ex. shigella variants)
- non-contagious diseases (ex. pufferfish neurotoxin)
- drop n/as and unkown diseases-- we can bring them back later but for now get rid of em

for the vector
- summary stats on cleaned cleaned data for where people r getting sick

Etiology variable cleaning loop
notes:
- dropped non-pathogenic observations (contaminants)
- reduced diseases to genus rather than specific species
- dropped unknowns and blank observations
```{r}
#before picking out which etiologies to exclude we will label things in prep

#for each row in the refined dataframe...
for (i in 1:nrow(refined)){ #REMEMBER i in 1:nrow(refined)
  string = refined$Etiology[i]
  #save whatever is in that row's Etiology column into a string
  
  #we will be splitting these into multiple individual dataframes to deal with individually
  #for reference: this df uses commas to specify pathogen variants and semicolons to separate multiple pathogens
  
  if (grepl(";", string)==TRUE){ #IF there are multiple diseases listed...
    string = unlist(strsplit(string, split = ";"))
    string = trimws(string)
    if(length(unique(string))==1){
      #if all of the diseases are the same disease, overwrite with just one
      refined$Etiology[i] = toString(unique(string))
      refined$Etiology.Expanded[i] = toString(unique(string))
      }
    else{
      #if the diseases are actually different, put "multiple" in etiology field
      refined$Etiology[i] = "Multiple"
    }
  }
}

message("first loop done!")

#ok NOW we pull out the variables that we don't want
#pulling out things that are NOT pathogens
#and kept pathogens are being reduced to their genus

noncontagious = c("Alkaline Water", "Amnesic shellfish poison","Anatoxin-a", "Arsenic", "Bromine", "Chloramines", "Chlordane", "Chlorine", "Chlorine Gas", "Chromium", "Ciguatoxin", "Cleaning agents", "Copper", "Ethylene Glycol", "Fluoride", "Gasoline", "Heavy metals", "Histamine", "Hydrochloric Acid", "Monosodium glutamate (MSG)", "Morpholine", "Neurotoxic shellfish poison","Nitrate","Nitrite", "Oil", "Other", "Other - Chemical/Toxin", "Paralytic shellfish poison", "Pesticides", "Phenol", "Plant/Herbal toxins", "Puffer fish tetrodotoxin", "Scombroid toxin", "Selenium", "Soap/Detergent", "Sodium Hydroxide", "Trichloroethylene", "Unknown - Chemical/Toxin", "ethyl acrylate", "herbicide", "Unknown", "")

contagious = c("Acanthamoeba", "Astrovirus", "Anisakis", "Bacillus", "Campylobacter", "Clostridium", "Cryptosporidium", "Cyclospora", "Entamoeba", "Enterobacter", "Enterococcus", "Enterovirus", "Escherichia", "Giardia", "Hepatovirus", "Legionella", "Leptospira", "Listeria", "Microcystis", "Mycobacterium", "Naegleria", "Norovirus", "Pantoea", "Plesiomonas", "Providencia", "Pseudomonas", "Rotavirus", "Salmonella", "Sapovirus", "Shigella", "Staphylococcus", "Streptococcus", "Vibrio")

unsure = c("Avian schistosomes", "Cyanotoxin", "Cyanotoxin other", "Cyanotoxin unknown", "Microcoleus lyngbyei", "Mycotoxins", "Other - Bacterium", "Other - Parasite", "Other - Virus", "Unknown - Bacterium", "Unknown - Virus")

drop2 = refined |> filter(Etiology %in% noncontagious | Etiology %in% unsure)
refined = refined |> filter(!(Etiology %in% noncontagious | Etiology %in% unsure))

for (i in 1:nrow(refined)){ #nrow(refined)
  string = trimws(refined$Etiology[i])
  if(string == "Multiple"){
    next
  }
  else if(grepl("Acanthamoeba", string)==TRUE){
    refined$Etiology[i] = "Acanthamoeba"
  }
  else if(grepl("Adenovirus", string)==TRUE){
    refined$Etiology[i] = "Adenovirus"
    #adenovirus is not a genus but only one of two genera in the group infect humans
  }
  else if(grepl("Bacillus", string)==TRUE){
    refined$Etiology[i] = "Bacillus"
  }
  else if(grepl("Brucella", string)==TRUE){
    refined$Etiology[i] = "Brucella"
  }
  else if(grepl("Campylobacter", string)==TRUE){
    refined$Etiology[i] = "Campylobacter"
  }
  else if(grepl("Clostridium", string)==TRUE){
    refined$Etiology[i] = "Clostridium"
  }
  else if(grepl("Cryptosporidium", string)==TRUE){
    refined$Etiology[i] = "Cryptosporidium"
  }
  else if(grepl("Cyclospora", string)==TRUE){
    refined$Etiology[i] = "Cyclospora"
  }
  else if(grepl("Entamoeba", string)==TRUE){
    refined$Etiology[i] = "Entamoeba"
  }
  else if(grepl("Enterococcus", string)==TRUE){
    refined$Etiology[i] = "Enterococcus"
  }
  else if(grepl("Enterovirus", string)==TRUE){
    refined$Etiology[i] = "Enterovirus"
  }
  else if(grepl("Escherichia", string)==TRUE){
    refined$Etiology[i] = "Escherichia"
  }
  else if(grepl("Giardia", string)==TRUE){
    refined$Etiology[i] = "Giardia"
  }
  else if(grepl("Hepatitis", string)==TRUE){
    refined$Etiology[i] = "Hepatovirus "
  }
  else if(grepl("Legionella", string)==TRUE){
    refined$Etiology[i] = "Legionella"
  }
  else if(grepl("Leptospira", string)==TRUE){
    refined$Etiology[i] = "Leptospira"
  }
  else if(grepl("Listeria", string)==TRUE){
    refined$Etiology[i] = "Listeria"
  }
  else if(grepl("Microcystis", string)==TRUE){
    refined$Etiology[i] = "Microcystis"
  }
  else if(grepl("Mycobacterium", string)==TRUE){
    refined$Etiology[i] = "Mycobacterium"
  }
  else if(grepl("Naegleria", string)==TRUE){
    refined$Etiology[i] = "Naegleria"
  }
  else if(grepl("Norovirus", string)==TRUE){
    refined$Etiology[i] = "Norovirus"
  }
  else if(grepl("Pantoea", string)==TRUE){
    refined$Etiology[i] = "Pantoea"
  }
  else if(grepl("Plesiomonas", string)==TRUE){
    refined$Etiology[i] = "Plesiomonas"
  }
  else if(grepl("Providencia", string)==TRUE){
    refined$Etiology[i] = "Providencia"
  }
  else if(grepl("Pseudomonas", string)==TRUE){
    refined$Etiology[i] = "Pseudomonas"
  }
  else if(grepl("Salmonella", string)==TRUE){
    refined$Etiology[i] = "Salmonella"
  }
  else if(grepl("Sapovirus", string)==TRUE){
    refined$Etiology[i] = "Sapovirus"
  }
  else if(grepl("Shigella", string)==TRUE){
    refined$Etiology[i] = "Shigella"
  }
  else if(grepl("Staphylococcus", string)==TRUE){
    refined$Etiology[i] = "Staphylococcus"
  }
  else if(grepl("Streptococcus", string)==TRUE){
    refined$Etiology[i] = "Streptococcus"
  }
  else if(grepl("Toxoplasma", string)==TRUE){
    refined$Etiology[i] = "Toxoplasma"
  }
  else if(grepl("Trichinella", string)==TRUE){
    refined$Etiology[i] = "Trichinella"
  }
  else if(grepl("Vibrio", string)==TRUE){
    refined$Etiology[i] = "Vibrio"
  }
  else if(grepl("Yersinia", string)==TRUE){
    refined$Etiology[i] = "Yersinia"
  }
  else{
    next
  }
}

dropped <- bind_rows(dropped, drop2)
rm(drop2)


message("all done!")

```


dropping non-state and non-continental territories + multiple etiology outbreaks
```{r}
refined = refined |> 
  filter(!(State %in% c("Alaska", "Multistate", "Puerto Rico", "Hawaii", "District of Columbia", "Northern Mariana Islands", "Guam", "Republic of Palau", "Virgin Islands", "Republic of the Marshall Islands"))) |> 
  filter(Etiology != "Multiple")
#loss of 1424 non-continental outbreaks
#and 61 outbreaks in DC
# plus 1360 outbreaks with multiple etiologies
dropping = refined |> filter(State %in% c("Alaska", "Multistate", "Puerto Rico", "Hawaii", "District of Columbia", "Northern Mariana Islands", "Guam", "Republic of Palau", "Virgin Islands", "Republic of the Marshall Islands")) |> 
  filter(Etiology != "Multiple")
dropped <- bind_rows(dropped, dropping)
rm(dropping)

```

rename and group states by NOAA region
```{r}
northwest = c("Oregon", "Washington", "Idaho")
west = c("California", "Nevada")
southwest = c("New Mexico", "Arizona", "Colorado", "Utah")
west_north_central = c("Montana", "Wyoming", "North Dakota", "South Dakota", "Nebraska")
east_north_central = c("Minnesota", "Wisconsin", "Michigan", "Iowa")
central = c("Missouri", "Illinois", "Indiana", "Ohio", "Kentucky", "Tennessee", "West Virginia")
south = c("Texas", "Oklahoma", "Kansas", "Arkansas", "Louisiana", "Mississippi")
southeast = c("Alabama", "Georgia", "Florida", "North Carolina", "South Carolina", "Virginia")
northeast = c("Maryland", "Delaware", "New Jersey", "Connecticut", "Rhode Island", "Pennsylvania", "New York", "Massachusetts", "New Hampshire", "Vermont", "Maine")

for (i in 1:nrow(refined)){
 if (refined$State[i] %in% northwest){
   refined$State[i] = "Northwest"
   }
  else if (refined$State[i] %in% west){
    refined$State[i] = "West" 
  }
  else if (refined$State[i] %in% southwest){
    refined$State[i] = "Southwest"
  }
  else if (refined$State[i] %in% west_north_central){
    refined$State[i] = "West North Central"
  }
  else if (refined$State[i] %in% east_north_central){
    refined$State[i] = "East North Central"
  }
  else if (refined$State[i] %in% central){
    refined$State[i] = "Central"
  }
  else if (refined$State[i] %in% south){
    refined$State[i] = "South"
  }
  else if (refined$State[i] %in% southeast){
    refined$State[i] = "Southeast"
  }
  else if (refined$State[i] %in% northeast){
    refined$State[i] = "Northeast"
  }
}

rm(i, northwest, west, southwest, west_north_central, east_north_central, central, south, southeast, northeast)

refined = rename(refined, Region = State)


```



save this and then next chunk will read it in so i don't need to run this every time
```{r}

write.csv(dropped, "data/NORS_dropped.csv", row.names = FALSE)
write.csv(refined, "data/NORS_refined.csv", row.names = FALSE)

```
