
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)

NORS_alltime = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/NORS_alltime.csv", header = TRUE)
NORS = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/NORS_refined.csv", header = TRUE)
NOAA = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/tempdata/temp_reg.csv", header = TRUE)

#keeps reading in the observation number as a column
#idk why but i'll just delete it manually
#NOAA$X <- NULL
#NORS$X <- NULL
```

NORS observation distributions (preliminary)
```{r}
#REPORTING BY YEAR
NORS_alltime |> group_by(Year) |> 
  summarize(n = n())

NORS_alltime |> group_by(Year) |> 
  summarize(n = n()) |> 
  ggplot(aes(Year))+
  geom_histogram(aes(weight = n), color = 'darkgrey', bins = 52)+
  scale_x_continuous(n.breaks = 11, minor_breaks = NULL, limits = c(1970, 2022))+
  geom_vline(xintercept = 1998, linetype = 'dashed', linewidth = 1, color = 'chartreuse3')+
  geom_vline(xintercept = c(1980, 2009, 2020), linetype = 'dashed', linewidth = 1, color = 'darkred')+
  ylab("Outbreak Count")
ggsave("./figures/histogram_full_labeled.png", height = 4.5, width = 7)
#no decent data until 1998
#data also ends at 2022 but we will cut it off at 2019 to avoid influence from the pandemic



# REPORTING BY YEAR - LIMITED TO 1998-2019
NORS |> group_by(Year) |> 
  summarize(n = n())
#NEW by year HISTOGRAM
NORS |> group_by(Year) |> 
  summarize(n = n()) |> 
  ggplot(aes(Year))+
  geom_histogram(aes(weight = n), color = 'darkgrey', bins = 22)+
  scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020), 
                     minor_breaks = NULL)+
  ylab("Outbreak Count")
ggsave("./figures/histogram_trimmed.png", height = 4.5, width = 7)
#the amount of reporting is still definitely skewed towards more recent years
#but we're up to an n that I'm not too worried about it?

# REPORTING BY NOAA REGION
NORS |> group_by(Region) |> 
  summarize(n = n())
NORS |> group_by(Region) |> 
  summarize(n = n()) |> 
  ggplot()+
  geom_bar(aes(x = Region, weight = n))+
  ylab("Outbreak Count")+
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
ggsave("./figures/histogram_region_trimmed.png", height = 4.5, width = 7)

#tbh i'm not even sure how to go about cleaning this
#what even are relevant categories to divide by?

# REPORTING BY ETIOLOGY
NORS |> count(Etiology) |> 
  arrange(desc(across(n)))

NORS |> 
  count(Etiology) |> 
  arrange(desc(across(n)))|>
  head(10) |> 
  ggplot(aes(x = Etiology, y = n))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 20, hjust = 1))+
  ylab("Outbreak Count")
ggsave("./figures/histogram_etiology_all.png", height = 4.5, width = 7)


NORS |> 
  count(Etiology) |> 
  arrange(desc(across(n)))|>
  head(10) |> 
  filter(Etiology != "Norovirus") |> 
  ggplot(aes(x = Etiology, y = n))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 20, hjust = 1))+
  ylab("Outbreak Count")
ggsave("./figures/histogram_etiology_sinNoro.png", height = 4.5, width = 7)


```



NOAA observations distributions
```{r}
#we've already cropped some of the older data from the NOAA dataset
#but let's be sure we're only looking at 1998-2019
#starting with 2646 observations in the original dataframe

NOAA_cropped = drop_na(NOAA, c("Oct", "Nov", "Dec"))
#this drops 98 entries (2548 left)
#HOWEVER: please note that just filtering by year also drops these entries
#so the NAs aren't even in our scope and we have full data for the entire timespan

NOAA_cropped = NOAA_cropped |> 
  filter(Year >= 1998 & Year <= 2019) |> 
  filter(StateID < 49)
#2156 entries left (dropped 490)
#limited stateids to the continental 48


NOAA_cropped$StateID <- as.factor(NOAA_cropped$StateID)

NOAA_cropped |> 
  group_by(StateID) |> 
  summarize(n = length(StateID))

```

run this to delete all objects except the most recent datasets
```{r}
#NORS variables
rm(list = c("NORS", "NORS_state", "NORS_year"))

#NOAA variables
rm(list = c("NOAA"))

```


The NOAA and the NORS data have now been restricted to the same time frame.
Moving onto summary stats
NOAA
- temperature range
  - change over the decades
  - variation within a year
- state coverage
  - regions?
  
NORS
- disease frequency over time
  - can we even do this with varying levels of de
  

NOAA summary stats
```{r}
#NOAA_cropped |>  
#  group_by(StateID) |> 
#  summarize(n = n())
#every region has n = 22 obs meaning that we have good coverage! yippee

NOAA_pivoted = NOAA_cropped |> 
  pivot_longer(
    cols = !(c(ID, StateID, Year)),
    names_to = "Month",
    values_to = "Temp"
  )
#had to pivot the data so that all of the temps are in one column 
#other columns tell me the corresponding month/year
NOAA_pivoted$Month = as.factor(NOAA_pivoted$Month)
#this puts the months in the correct order.
NOAA_pivoted$Month = fct_relevel(NOAA_pivoted$Month, c("Jan", "Feb", "Mar", "Apr", "May", "June", "July", "Aug", "Sept", "Oct", "Nov", "Dec"))


#this graph is useless LMFAO
#NOAA_pivoted |> 
#  ggplot(aes(x = Year, y = Temp, color = Month))+
#  geom_point()+
#  geom_jitter()

NOAA_pivoted |> 
  filter(StateID == 33) |> 
  ggplot(aes(x = Year, y = Temp, color = Month)) +
  geom_line()
# this graph is a little less useless-- type in a state and it'll show you the yearly temps

NOAA_pivoted |> 
  filter(StateID == 12) |> 
  ggplot(aes(x = Year, y = Temp, color = Month)) +
  geom_line()

NOAA_pivoted |> 
  filter(StateID == 32) |> 
  ggplot(aes(x = Year, y = Temp, color = Month)) +
  geom_line()



NOAA_pivoted$Year = as.factor(NOAA_pivoted$Year) 




NOAA_pivoted |> 
  filter(Year == 1998 | Year == 2019) |> 
  group_by(Year) |> 
  ggplot(aes(x = Month, y = Temp, fill = Year))+
  geom_boxplot()

  #colder winters warmer summers?
#more variation in more recent years

```

NORS summary stats
```{r}
#NORS_cropped |> 
#  ggplot(aes(x = Year, color = Etiology))+
#  geom_line(y = )

NORS_counts = NORS_cropped |> 
  group_by(Year, Etiology) |> 
  summarize(n = n())

NORS_counts |> 
  filter(Etiology %in% topetiology$Etiology) |> 
  ggplot(aes(x = Year, y = n, color = Etiology))+
  geom_line()


```

```{r}

yearcounts = NORS_cropped |> 
  group_by(Year) |> 
  summarize(n = n()) |> 
  arrange(desc(n))

topetiology = NORS_cropped |> 
  group_by(Etiology) |> 
  summarize(n = n()) |> 
  arrange(desc(n)) |> 
  top_n(6)
  


```



exploring lists (and counts) of each unique etiology entry to prep for pruning
```{r}





single_factors = refined |>  
  filter(!is.na(Etiology)) |> 
  filter(Etiology != "Multiple")|> 
  group_by(Etiology) |> 
  summarize(n = n())

single_factors
#listed_factors = refined |> 
#  filter(Etiology == "Multiple") |> 
#  group_by(Etiology.Expanded) |> 
#  summarize(n = n())

#write.csv(single_factors, "etiologies_single.csv")
#write.csv(listed_factors, "listed_factors.csv")

```


To Do:
- WRITE
- normalize outbreaks to reporting? 
  - convert counts to yearly proportions?
- finish reading the tmap documentation & create maps for temp across the time
- keep only state data(?)
- 



