```{r}
library(dplyr)
library(tidyverse)
#library(forecast)
library(tseries)
library(ggplot2)
#library(ggplotify)
library(knitr)
library(extrafont)
library(viridis)
```

load in data
```{r}
temper = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/tempdata/temp_reg.csv", header = TRUE)
precip = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/precipdata/hum_reg.csv", header = TRUE)
NORS = read.csv("/Users/machaango/OneDrive/Desktop/thesis/zapata_thesis/data/NORS_refined.csv", header = TRUE)

regions_list = c("Central","East North Central", "Northeast", "Northwest", "South", "Southeast", "Southwest", "West", "West North Central")

```

visual theme settings (run only once)
```{r}
theme = theme_set(theme_linedraw())
theme_update(
  text=element_text(size=12,  family="Palatino Linotype"),
  panel.grid.major = element_line(
    size = 0.5, 
    linetype = 'solid',
    colour = "grey"),
  panel.grid.minor = element_line(
    size = 0.5, 
    linetype = 'solid',
    colour = "grey"))
```

custom functions
```{r}
numMonth = function(df){
  for (i in 1:nrow(df)){ 
    temp = df$Month[i]
    if(temp == "Jan"){
      df$Month[i] = 1
      }
    else if(temp == "Feb"){
      df$Month[i] = 2
      }
    else if(temp == "Mar"){
      df$Month[i] = 3
      }
    else if(temp == "Apr"){
      df$Month[i] = 4
      }
    else if(temp == "May"){
      df$Month[i] = 5
      }
    else if(temp == "June"){
      df$Month[i] = 6
      }
    else if(temp == "July"){
      df$Month[i] = 7
      }
    else if(temp == "Aug"){
      df$Month[i] = 8
      }
    else if(temp == "Sept"){
      df$Month[i] = 9
      }
    else if(temp == "Oct"){
      df$Month[i] = 10
      }
    else if(temp == "Nov"){
      df$Month[i] = 11
      }
    else if(temp == "Dec"){
      df$Month[i] = 12
      }
    }
  
  rm(i, temp)
  df$Month = as.integer(df$Month)
  df$Region = as.factor(df$Region)
  
  df = df |> mutate(Date2 = as.Date(paste(Year, Month, "01", sep = "-")))

  return(df)
  }


slopefxn = function(slope, intercept, val){
  out = (slope*val+intercept)
  return(out)
}

subtractingfxn = function(list1, list2){
  return(mapply("-", list1, list2))
}



```

pivot and vectorization (change variable based on need)
```{r}
target = "Temperature"
if (target == "Precipitation"){
  piv = precip
} else if (target == "Temperature"){
  piv = temper
} else {
  stop("change top line of code!")
  }

piv = piv |>
  pivot_longer(
    cols = !(c(ID, Region, Year)),
    names_to = "Month",
    values_to = target
  ) |> 
  select(!("ID"))
piv = numMonth(piv)

if (target == "Precipitation"){
  vec = piv |> 
  group_by(Region) |> 
  summarize(
    Precipitation = list(Precipitation))
  precip_piv = piv
  precip_vec = vec
} else if (target == "Temperature"){
  vec = piv |> 
  group_by(Region) |> 
  summarize(
    Temperature = list(Temperature)) 
  temp_piv = piv
  temp_vec = vec
}

rm(piv, vec, target)
```

deconstructing I: are our time series stationary?
```{r}
for (i in 1:nrow(temp_vec)){
  temp = unlist(temp_vec$Temperature[i])
  temp_vec$ADF_pval[i] = adf.test(temp, k = 12)$p.value
}
#yes! stationary

for (i in 1:nrow(precip_vec)){
  temp = unlist(precip_vec$Precipitation[i])
  precip_vec$ADF_pval[i] = adf.test(temp, k = 12)$p.value
}
#not stationary?

temp = temp_vec |> select(Region, ADF_pval)
tibble(temp)
temp = precip_vec |> select(Region, ADF_pval)
tibble(temp)
rm(temp)

```



deconstructing II: visualizing the time series decomp (table)
```{r}
# path names for when these are saved
regions_paths_precip = c(
  "./figures/timeseries_decomp/precip_decomp_Central.png",
  "./figures/timeseries_decomp/precip_decomp_EastNorthCentral.png", 
  "./figures/timeseries_decomp/precip_decomp_Northeast.png",
  "./figures/timeseries_decomp/precip_decomp_Northwest.png",
  "./figures/timeseries_decomp/precip_decomp_South.png",
  "./figures/timeseries_decomp/precip_decomp_Southeast.png",
  "./figures/timeseries_decomp/precip_decomp_Southwest.png",
  "./figures/timeseries_decomp/precip_decomp_West.png",
  "./figures/timeseries_decomp/precip_decomp_WestNorthCentral.png")

regions_paths_temper = c(
  "./figures/timeseries_decomp/temper_decomp_Central.png",
  "./figures/timeseries_decomp/temper_decomp_EastNorthCentral.png", 
  "./figures/timeseries_decomp/temper_decomp_Northeast.png",
  "./figures/timeseries_decomp/temper_decomp_Northwest.png",
  "./figures/timeseries_decomp/temper_decomp_South.png",
  "./figures/timeseries_decomp/temper_decomp_Southeast.png",
  "./figures/timeseries_decomp/temper_decomp_Southwest.png",
  "./figures/timeseries_decomp/temper_decomp_West.png",
  "./figures/timeseries_decomp/temper_decomp_WestNorthCentral.png")


# plotting and saving precipitation data
for (i in 1:nrow(precip_vec)){
  #doing the decomposition
  temp = unlist(precip_vec$Precipitation[i])
  tempts = ts(temp, frequency = 12, start = 1998)
  saveplot = decompose(tempts)
  
  #plot label prep
  plot_subtitle = paste(regions_list[i], "Region Precipitation Data")

  #making the plot
  autoplot(saveplot, range.bars = FALSE)+
    scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020))+
    scale_y_continuous(minor_breaks = NULL)+
    xlab("Year")+
    labs(title = "Time Series Decomposition",
         subtitle = plot_subtitle)
  ggsave(regions_paths_precip[i], height = 5, width = 7)
  
  #no! not stationary. we have a trend :D
  #additionally, these are the time series based on the periodicity of the temperature data
  #however it looks as if there's extra autocorrelation that i'm not accounting for-- el Nino and other meteorological stuff 
}


# and doing the same for the temperature data
for (i in 1:nrow(temp_vec)){
  #doing the decomposition
  temp = unlist(temp_vec$Temperature[i])
  tempts = ts(temp, frequency = 12, start = 1998)
  saveplot = decompose(tempts)
  
  #plot label prep
  plot_subtitle = paste(regions_list[i], "Region Temperature Data")

  #making the plot
  autoplot(saveplot, range.bars = FALSE)+
    scale_x_continuous(breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020))+
    scale_y_continuous(minor_breaks = NULL)+
    xlab("Year")+
    labs(title = "Time Series Decomposition",
         subtitle = plot_subtitle)
  ggsave(regions_paths_temper[i], height = 5, width = 7)
}


#NOTES
#all temp are nonsignificant (stationary)
#all Precipitation are somewhat significant (NOT stationary)
#except for West North Central Region which is the most landlocked?

# clean up the environment
rm(regions_paths_precip, regions_paths_temper, temp, tempts, saveplot, plot_subtitle, i)
```

raw timeseries by region (plot)
```{r}
precip_signif = precip_piv |> filter(Region != "West North Central")

precip_piv |> 
  ggplot(aes(x = Date2, y = Precipitation))+
  geom_line()+
  facet_wrap(vars(Region))+
  geom_smooth(data = precip_signif, method = lm, color = 'red3')+
  xlab("Year")+
  labs(title = "Monthly Precipitation (in) by Region")
ggsave("./figures/precip_region.png", height = 4, width = 8)

temp_piv |> 
  group_by(Region) |> 
  mutate(avg_temp = mean(Temperature)) |> 
  ggplot(aes(x = Date2, y = Temperature))+
  geom_line()+
  #geom_segment(aes(x = as.Date('1998-01-01'), xend = as.Date('2019-12-01'), y = avg_temp, yend = avg_temp), color = 'blue')+
  facet_wrap(vars(Region))+
  xlab("Year")+
  labs(title = "Monthly Temperature (°F) by Region")
ggsave("./figures/temper_region.png", height = 4, width = 8)

rm(precip_signif)
```

extracting average value for all ts and slope+intercept for significant ts (table)
```{r}
temp_vec = temp_vec |> mutate(Avg = 0)
precip_vec = precip_vec |> mutate(Avg = 0,
                                  Slope = 0,
                                  Intercept = 0,
                                  Slope_P = 0)

for (i in 1:nrow(temp_vec)){
  temp = unlist(temp_vec$Temperature[i])
  temp_vec$Avg[i] = mean(temp)
  }

for (i in 1:nrow(precip_vec)){
  temp = unlist(precip_vec$Precipitation[i])
  precip_vec$Avg[i] = mean(temp)

  if (precip_vec$Region[i] != "West North Central"){
    precip_vec$Slope[i] = coef(model)[2]
    precip_vec$Intercept[i] = coef(model)[1]
    precip_vec$Pval[i] = summary(model)$coefficients[2,4] 
    }
}

temp = temp_vec |> select(!(c(ADF_pval, Temperature)))
tibble(temp)
temp = precip_vec |> select(!(c(ADF_pval, Precipitation)))
tibble(temp)

rm(temp)
```

trying to save autocorrelation plots
```{r}
for (i in 1:nrow(precip_vec)){
  temp = unlist(precip_vec$Precipitation[i])
  acf(temp)
  #pacf(temp, lag = 12)
  #adf.test(temp, k = 48)
  #precip_vec$adf.p[i] = adf.test(temp, k = 12)$p.value
}



for (i in 1:nrow(precip_vec)){
  temp = unlist(precip_vec$Precipitation[i])
  acf_saved = acf(temp, plot = FALSE)
  plot(acf_saved)
  acf_df <- with(acf_saved, data.frame(lag, acf))

  acf_df |> ggplot(mapping = aes(x = lag, y = acf))+
    geom_hline(aes(yintercept = 0))+
    geom_segment(mapping = aes(xend = lag, yend = 0))
  #pacf(temp, lag = 12)
  #adf.test(temp, k = 48)
  #precip_vec$adf.p[i] = adf.test(temp, k = 12)$p.value
}

bacf <- acf(x, plot = FALSE)
bacfdf <- with(bacf, data.frame(lag, acf))

q <- ggplot(data = bacfdf, mapping = aes(x = lag, y = acf)) +
       geom_hline(aes(yintercept = 0)) +
       geom_segment(mapping = aes(xend = lag, yend = 0))




```



tried adjusting the non-stationary time series to be stationary (non-functional script)
```{r}
precip_vec = precip_vec |> mutate(
  Intercept = 0,
  Slope = 0,
  SE = 0, 
  Predicted = 0,
  Adjusted = 0
)

for (i in 1:nrow(precip_vec)){
  temp_df = precip_piv |> filter(Region == regions_list[i])
  model = lm(Precipitation ~ Date2, data = temp_df)

  precip_vec$Intercept[i] = model$coef[[1]]
  precip_vec$Slope[i] = model$coef[[2]]
  precip_vec$SE = summary(model)$coef[[4]] 
  
  temp = unlist(precip_vec$Precipitation[i])
  precip_vec$Predicted[i] = list(lapply(temp, slopefxn, model$coef[[2]], model$coef[[1]]))
  
  temp2 = unlist(precip_vec$Predicted[i])
  precip_vec$Adjusted[i] = list(mapply("-", temp, temp2))
}

precip_piv = precip_piv |> 
  mutate(Predicted = 0)

for (i in 1:nrow(precip_piv)){
  if (precip_piv$Region[i] == "Central"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[1], precip_vec$Intercept[1], precip_piv$Precipitation[i])
  }
  else if (precip_piv$Region[i] == "East North Central"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[2], precip_vec$Intercept[2], precip_piv$Precipitation[i])
  } 
  else if (precip_piv$Region[i] == "Northeast"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[3], precip_vec$Intercept[3], precip_piv$Precipitation[i])
  } 
  else if (precip_piv$Region[i] == "Northwest"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[4], precip_vec$Intercept[4], precip_piv$Precipitation[i])
  } 
  else if (precip_piv$Region[i] == "South"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[5], precip_vec$Intercept[5], precip_piv$Precipitation[i])
  } 
  else if (precip_piv$Region[i] == "Southeast"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[6], precip_vec$Intercept[6], precip_piv$Precipitation[i])
  } 
  else if (precip_piv$Region[i] == "Southwest"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[7], precip_vec$Intercept[7], precip_piv$Precipitation[i])
  } 
  else if (precip_piv$Region[i] == "West"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[8], precip_vec$Intercept[8], precip_piv$Precipitation[i])
  } 
  else if (precip_piv$Region[i] == "West North Central"){
    precip_piv$Predicted[i] = slopefxn(precip_vec$Slope[9], precip_vec$Intercept[9], precip_piv$Precipitation[i])
  } 
  else{
    stop("idk what happened")
  }
}

precip_piv = precip_piv |> mutate(
  Adjusted = Precipitation-Predicted
)

precip_piv |> 
  ggplot(aes(x = Date2, y = Precipitation))+
  geom_line()+
  geom_line(aes(y = Adjusted), color = "blue")+
  facet_wrap(vars(Region))+
  geom_smooth(method = lm, color = "red")

precip_piv |> 
  ggplot(aes(x = Date2, y = Precipitation))+
  geom_line()+
  geom_line(aes(y = Adjusted), color = "blue")+
  facet_wrap(vars(Region))+
  geom_smooth(method = lm, color = "red")

#not entirely sure I did this right
#regardless, the point of this was to extract and look at the slopes (which I have)
#i tried making it stationary so we can rejoin the temp dataset but alas.

```


chi test stuff + graph
```{r}
low_counts = NORS |> group_by(Etiology) |> count() |> arrange(n) |> filter(n < 15)
low_counts = low_counts$Etiology

chitest1 = NORS |> 
  group_by(Region, Etiology) |> 
  count() |> 
  pivot_wider(names_from = Region, values_from = n, values_fill = 0)

chitest2 = chitest1 |> ungroup() |> select(!(Etiology))
#https://stats.stackexchange.com/questions/159057/alternatives-for-chi-squared-test-for-independence-for-tables-more-than-2-x-2
chisq.test(chitest2, simulate.p.value = TRUE) #the nature of this test means that it's testing the RELATIVE proportion-- it assumes that the spread overall is what matters and not differing expected counts between groups
#i think that's what that means?
#regardless it's actually running a fisher's exact test to account for low values

## VISUALS
## total proportion
NORS |> filter(!(Etiology %in% low_counts | Etiology == "Norovirus")) |> ggplot(aes(x = Region, fill = Etiology))+
  geom_bar(position = "stack")

# relative proportion
NORS |> filter(!(Etiology %in% low_counts|Etiology == "Norovirus")) |> ggplot(aes(x = Region, fill = Etiology))+
  geom_bar(position = "fill")

rm(low_counts, chitest1, chitest2)
```

checking hot/cold months
```{r}
temp_piv |> 
  ggplot(aes(x = Month, y = Temperature, color = Year))+
  geom_point(alpha = 0.5)+
  scale_x_continuous(breaks=seq(1,12,1))+
  facet_wrap(vars(Region))+
  labs(title = "Monthly Temperature (°F) by Region")+
  scale_color_viridis(option="viridis")
ggsave("./figures/temp_monthly.png")
# hottest months: june, july, aug
# coldest months: jan, feb, dec

precip_piv |> 
  ggplot(aes(x = Month, y = Precipitation, color = Year))+
  geom_point(alpha = 0.5)+
  scale_x_continuous(breaks=seq(1,12,1))+
  facet_wrap(vars(Region))+
  labs(title = "Monthly Precipitation (inches) by Region")+
  scale_color_viridis(option="viridis")
ggsave("./figures/precip_monthly.png")

  
#each region has its own unique Precipitation pattern
```

prepping NORS for proportion comparisons
```{r}
#NORS |> filter(!(Etiology %in% low_counts | Etiology == "Norovirus")) |> group_by(Etiology) |> count()
#NORS|> group_by(Etiology) |> count()
low_counts = NORS |> group_by(Etiology) |> count() |> arrange(n) |> filter(n < 15)
low_counts = low_counts$Etiology

NORS2 = NORS |> 
  select(c(Year, Month, Region, Etiology)) |> 
  mutate(Type = "low count",
         Special_Temp = "none", 
         Section = 0)
NORS2 = numMonth(NORS2)
NORS2 = NORS2 |> rename(Date = Date2)

for (i in 1:nrow(NORS2)){
  if (NORS2$Month[i] <= 8 & NORS2$Month[i] >= 6){
    NORS2$Special[i] =  "hot"
  }
  if (NORS2$Month[i] <= 2 | NORS2$Month[i] == 12){
    NORS2$Special[i] = "cold"
  }
}

for (i in 1:nrow(NORS2)){
  if (NORS2$Month[i] <= 8 & NORS2$Month[i] >= 6){
    NORS2$Special[i] =  "hot"
  }
  if (NORS2$Month[i] <= 2 | NORS2$Month[i] == 12){
    NORS2$Special[i] = "cold"
  }
}


for (i in 1:nrow(NORS2)){
  if (NORS2$Etiology[i] == "Acanthamoeba"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Adenovirus"){
    NORS2$Type[i] = "Virus"
  }
  else if (NORS2$Etiology[i] == "Anisakis"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Astrovirus"){
    NORS2$Type[i] = "Virus"
  }
  else if (NORS2$Etiology[i] == "Bacillus"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Brucella"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Campylobacter"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Clostridium"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Cryptosporidium"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Cyclospora"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Enterobacter"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Enterococcus"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Escherichia"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Giardia"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Hepatovirus "){
    NORS2$Type[i] = "Virus"
  }
  else if (NORS2$Etiology[i] == "Legionella"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Leptospira"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Listeria"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Microcystis"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Mycobacterium"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Naegleria"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Norovirus"){
    NORS2$Type[i] = "Virus"
  }
  else if (NORS2$Etiology[i] == "Pantoea"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Plesiomonas"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Providencia"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Pseudomonas"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Rotavirus"){
    NORS2$Type[i] = "Virus"
  }
  else if (NORS2$Etiology[i] == "Salmonella"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Sapovirus"){
    NORS2$Type[i] = "Virus"
  }
  else if (NORS2$Etiology[i] == "Shigella"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Staphylococcus"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Streptococcus"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Toxoplasma"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Trichinella"){
    NORS2$Type[i] = "Parasite/Amoeba"
  }
  else if (NORS2$Etiology[i] == "Vibrio"){
    NORS2$Type[i] = "Bacteria"
  }
  else if (NORS2$Etiology[i] == "Yersinia"){
    NORS2$Type[i] = "Bacteria"
  }
  else{
    message(NORS2$Etiology[i])
  }
}


NORS_lite = NORS2 |> filter(!(Etiology %in% low_counts | Etiology == "Norovirus"))


message("all done!")
```


proportions by year and region
```{r}
#example of formatted stuff
temp_piv |> 
  ggplot(aes(x = Month, y = Temperature, color = Year))+
  geom_point(alpha = 0.5)+
  scale_x_continuous(breaks=seq(1,12,1))+
  facet_wrap(vars(Region))+
  labs(title = "Monthly Temperature (°F) by Region")+
  scale_color_viridis(option="viridis")
ggsave("./figures/temp_monthly.png")
# in this case we can't drop the loest etiologies because of how the disease is working-- many different types of virus make up the cold season but most hot month infections are caused by the same pathogen
# also. several amoeba groups are like. sort of parasitic sort of not (parasite has its own definition) but they have been grouped here 

# HOT/COLD x TYPE
NORS2 |> filter(Special != "none") |> 
  ggplot(aes(x = Special, fill = Type))+
  geom_bar(position = "stack")+
  facet_wrap(vars(Region))

NORS_lite |> filter(Special != "none") |> 
  ggplot(aes(x = Special, fill = Type))+
  geom_bar(position = "stack")+
  facet_wrap(vars(Region))


# HOT/COLD x ETIOLOGY (stacked)
NORS2 |> filter(Special != "none") |> 
  ggplot(aes(x = Special, fill = Etiology))+
  geom_bar(position = "stack")+
  facet_wrap(vars(Region))

NORS_lite |> filter(Special != "none") |> 
  ggplot(aes(x = Special, fill = Etiology))+
  geom_bar(position = "stack")+
  facet_wrap(vars(Region))

# HOT/COLD x ETIOLOGY (proportion)
NORS2 |> filter(Special != "none") |> 
  ggplot(aes(x = Special, fill = Etiology))+
  geom_bar(position = "fill")+
  facet_wrap(vars(Region))

NORS_lite |> filter(Special != "none") |> 
  ggplot(aes(x = Special, fill = Etiology))+
  geom_bar(position = "fill")+
  facet_wrap(vars(Region))




NORS2 |> 
  ggplot(aes(x = Month, fill = Type))+
  geom_bar(position = "stack")+
  facet_wrap(vars(Region))+
  ggtitle("Type x Month by Region (full)")

NORS_lite |> 
  ggplot(aes(x = Month, fill = Type))+
  geom_bar(position = "stack")+
  facet_wrap(vars(Region))+
  ggtitle("Type x Month by Region (lite)")

temp_piv |> 
  ggplot(aes(x = Month, y = Temperature, color = Year))+
  geom_point()+
  facet_wrap(vars(Region))+
  ggtitle("Temperature x Month by Region")
# hottest months: june, july, aug
# coldest months: jan, feb, dec

precip_piv |> 
  ggplot(aes(x = Month, y = Precipitation, color = Year))+
  geom_point()+
  facet_wrap(vars(Region))+
  ggtitle("Precipitation x Month by Region")+
  theme(text=element_text(size=12,  family="Palatino Linotype"))


```


```{r}
NORS_props = NORS2 |> group_by(Year, Region, Etiology) |> add_count()
NORS_props = NORS_props |> 
  ungroup() |> 
  group_by(Year, Region) |> 
  add_count(wt = n) |> 
  rename(indiv_count = n,
         total_count = nn) |> 
  mutate(prop = indiv_count/total_count)




# Precipitation etiology -- disease
```


